Estimating Predictive Rate-Distortion Curves via Neural Variational Inference

### 2019-06-16 paper.tex
Updates from Overleaf

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Stanford University; mhahn2@stanford.edu\\
$^{2}$ \quad UC Irvine; rfutrell@uci.edu}

@@

The problem of predictive rate-distortion has been formalized by a range of studies.
A principled and general formalization is provided by applying the Information Bottleneck idea \citep{tishby-information-1999, still-optimal-2010,marzen-predictive-2016}:
We will write $\past$ for the past $X_{< 0}$, and $\future$ for the future $X_{\geq 0}$, following \cite{marzen-predictive-2016}.
We consider random variables $Z$, called \key{codes}, that summarize the past and are used by the observer to predict the future.
Formally, $Z$ needs to be independent of the future $\future$ conditional on the past $\past$: in other words, $Z$ does not provide any information about the future except what is contained in the past. Symbolically:
\begin{equation}\label{eq:markov}
	Z \bot \future | \past.
\end{equation}
This is equivalent to the requirement that $Z \rightarrow \past \rightarrow \future$ be a Markov chain.
This formalizes the idea that the code is computed by an observer from the past, without having access to the future.
Predictions are then made based only on $Z$, without additional access to the past $\past$.

@@

The distortion is the loss in predictive accuracy when predicting from $Z$, relative to optimal prediction from the full past $\past$.
In the Information Bottleneck formalization, this is equivalent to the amount of mutual information between past and future that is \emph{not} captured by $Z$ \citep{harremoes2007information}: 
\begin{equation}
	\operatorname{I}[\past, \future|Z].
\end{equation}
Due to the Markov condition, the distortion measure satisfies the relation
\begin{equation}
	\operatorname{I}[\past, \future|Z] = \operatorname{I}[\past, \future] - \operatorname{I}[Z, \future],
\end{equation}
i.e., it captures how much less information $Z$ carries about the future $\future$ compared to the full past $\past$.
For a fixed process $(X_t)_t$, choosing $Z$ to minimize the distortion is equivalent to maximizing the mutual information between the code and the future:
\begin{equation}\label{eq:predictiveness}
\operatorname{I}[Z, \future]
\end{equation}
We will refer to~(\ref{eq:predictiveness}) as the \key{predictiveness} of the code $Z$.

@@

\begin{corollary}[Main Result]
The predictive rate-distortion objective~(\ref{eq:prd-loss})
\begin{equation}\tag{\ref{eq:prd-loss}}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \operatorname{I}[ \finitepast, Z]\right]
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q} \left[\mathbb{E}_{\finitepast, \finitefuture}	\left[-	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] + \lambda \cdot \operatorname{D_{KL}} [ \infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]\right],
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
where $\phi, \psi, q$ range over all triples of the appropriate types described above.

@@

It is important to note that this sampling error is different from the overfitting issue affecting OCF: 
(\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}) provide \emph{unbiased} estimators of upper bounds, whereas overfitting \emph{biases} the values obtained by OCF.
Given that NPRD provably provides upper bounds on rate and prediction loss (up to sampling error), one can objectively compare the quality of different estimation methods:
Among methods that provide upper bounds, the one that provides the lowest such bound for a given problem is the one giving results closest to the true curve.

Given the estimate for prediction loss, we estimate predictiveness $\operatorname{I}[Z, \finitefuture]$ with the following method.
%by estimating $\operatorname{H}[\finitefuture]$ 
We use the encoder network that computes the code vector $h$ (\ref{eq:z-gauss}) to also estimate the marginal probability of the past observation sequence, $P_\eta(\finitepast)$. Because we consider stationary processes, we have 
$- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitefuture)\right] = - \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitepast)\right]$.
Using this observation, we estimate the predictiveness $\operatorname{I}[Z, \finitefuture] = \operatorname{H}[\finitefuture] - \operatorname{H}[\finitefuture|Z]$ by the difference between the corresponding cross-entropies on the test set~\cite{mcallester2018formal}:
%$\operatorname{H}[\finitefuture]$ by the cross-entropy of $P_\eta$ on the test data:
\begin{equation}\label{eq:bound-mc-predictiveness}
- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitefuture) - \log P_\psi(\finitefuture | Z)\right],
\end{equation}
which we approximate using Monte-Carlo sampling on the test set as in~(\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}).

During optimization, we add the cross-entropy term $- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitepast)\right]$ to the PRD objective~(\ref{eq:bound}), so that the full training objective comes out to:
\begin{equation}\label{eq:bound-augmented}
	\min_{\phi, \psi, q,\eta} \left[\mathbb{E}_{\finitepast, \finitefuture}	\log P_\eta(\finitepast) + \left[-	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] + \lambda \cdot \operatorname{D_{KL}} [ \infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]\right].
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
Again by Gibbs' inequality and Propositions~\ref{prop:mi-bound} and~\ref{prop:kl-bound}, this is minimized when $P_\eta$ represents the true distribution over length-$M$ blocks $P(\finitepast)$, $P_\phi(Z|\finitepast)$ describes an optimal code for the given $\lambda$, $q$ is its marginal distribution, and $P_\psi(\finitefuture|Z)$ is the Bayes-optimal decoder.
For approximate solutions to this augmented objective, the inequalities~(\ref{eq:bound1-new}-\ref{eq:bound2-new}) will also remain true due to Propositions~\ref{prop:mi-bound} and~\ref{prop:kl-bound}.

@@

In the neural networks literature, the most commonly used method using variational bounds similar to (\ref{eq:bound}) is the Variational Autoencoder \citep{kingma-auto-encoding-2014, bowman-generating-2016}, which corresponds to the setting where $\lambda=1$ and the predicted output is equal to the observed input.
The $\beta$-VAE \citep{higgins2017beta}, a variant of the Variational Autoencoder, uses $\lambda > 1$ (whereas the Predictive Rate-Distortion objective~(\ref{eq:ib}) uses $\lambda \in (0,1)$), and has been linked to the Information Bottleneck by \cite{burgess2018understanding}.

@@

\section{Estimating PRD for Natural Language}\label{sec:language}

>>>>

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Stanford University; mhahn2@stanford.edu\\
$^{2}$ \quad University of California, Irvine; rfutrell@uci.edu}

@@

The problem of predictive rate-distortion has been formalized by a range of studies.
A principled and general formalization is provided by applying the Information Bottleneck idea \citep{tishby-information-1999, still-optimal-2010,marzen-predictive-2016}:
We will write $\past$ for the past $X_{< 0}$, and $\future$ for the future $X_{\geq 0}$, following \cite{marzen-predictive-2016}.
We consider random variables $Z$, called \key{codes}, that summarize the past and are used by the observer to predict the future.
Formally, $Z$ needs to be independent of the future $\future$ conditional on the past $\past$: in other words, $Z$ does not provide any information about the future except what is contained in the past. Symbolically:
\begin{equation}\label{eq:markov}
	Z \bot \future | \past.
\end{equation}
This is equivalent to the requirement that $Z \leftrightarrow \past \leftrightarrow \future$ be a Markov chain.
This formalizes the idea that the code is computed by an observer from the past, without having access to the future.
Predictions are then made based only on $Z$, without additional access to the past $\past$.

@@

The distortion is the loss in predictive accuracy when predicting from $Z$, relative to optimal prediction from the full past $\past$.
In the Information Bottleneck formalization, this is equivalent to the amount of mutual information between past and future that is \emph{not} captured by $Z$ \citep{harremoes2007information}: 
\begin{equation}
	\operatorname{I}[\past, \future|Z].
\end{equation}
Due to the Markov condition, the distortion measure satisfies the relation
\begin{equation}
	\operatorname{I}[\past, \future|Z] = \operatorname{I}[\past, \future] - \operatorname{I}[Z, \future],
\end{equation}
i.e., it captures how much less information $Z$ carries about the future $\future$ compared to the full past $\past$.
For a fixed process $(X_t)_t$, choosing $Z$ to minimize the distortion is equivalent to maximizing the mutual information between the code and the future:
\begin{equation}\label{eq:predictiveness}
\operatorname{I}[Z, \future].
\end{equation}
We will refer to~(\ref{eq:predictiveness}) as the \key{predictiveness} of the code $Z$.

@@

\begin{corollary}[Main Result]
The predictive rate-distortion objective~(\ref{eq:prd-loss})
\begin{equation}\tag{\ref{eq:prd-loss}}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \operatorname{I}[ \finitepast, Z]\right]
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q} \left[\mathbb{E}_{\finitepast, \finitefuture}	\left[-	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] + \lambda \cdot \operatorname{D_{KL}} \left[ \infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]\right]\right],
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
where $\phi, \psi, q$ range over all triples of the appropriate types described above.

@@

It is important to note that this sampling error is different from the overfitting issue affecting OCF: 
Our Equations (\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}) provide \emph{unbiased} estimators of upper bounds, whereas overfitting \emph{biases} the values obtained by OCF.
Given that NPRD provably provides upper bounds on rate and prediction loss (up to sampling error), one can objectively compare the quality of different estimation methods:
Among methods that provide upper bounds, the one that provides the lowest such bound for a given problem is the one giving results closest to the true curve.

Given the estimate for prediction loss, we estimate predictiveness $\operatorname{I}[Z, \finitefuture]$ with the following method.
%by estimating $\operatorname{H}[\finitefuture]$ 
We use the encoder network that computes the code vector $h$ (\ref{eq:z-gauss}) to also estimate the marginal probability of the past observation sequence, $P_\eta(\finitepast)$ \rljf{How?}. Because we consider stationary processes, we have 
$- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitefuture)\right] = - \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitepast)\right]$.
Using this observation, we estimate the predictiveness $\operatorname{I}[Z, \finitefuture] = \operatorname{H}[\finitefuture] - \operatorname{H}[\finitefuture|Z]$ by the difference between the corresponding cross-entropies on the test set~\cite{mcallester2018formal}:
%$\operatorname{H}[\finitefuture]$ by the cross-entropy of $P_\eta$ on the test data:
\begin{equation}\label{eq:bound-mc-predictiveness}
- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitefuture) - \log P_\psi(\finitefuture | Z)\right],
\end{equation}
which we approximate using Monte-Carlo sampling on the test set as in~(\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}).

During optimization, we add the cross-entropy term $- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitepast)\right]$ to the PRD objective~(\ref{eq:bound}), so that the full training objective comes out to:
\begin{equation}\label{eq:bound-augmented}
	\min_{\phi, \psi, q,\eta} \left[\mathbb{E}_{\finitepast, \finitefuture} \left[ -	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] + \lambda \cdot \operatorname{D_{KL}} \left[ \infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right] - \log P_\eta(\finitepast) \right]\right].
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
Again by Gibbs' inequality and Propositions~\ref{prop:mi-bound} and~\ref{prop:kl-bound}, this is minimized when $P_\eta$ represents the true distribution over length-$M$ blocks $P(\finitepast)$, $P_\phi(Z|\finitepast)$ describes an optimal code for the given $\lambda$, $q$ is its marginal distribution, and $P_\psi(\finitefuture|Z)$ is the Bayes-optimal decoder.
For approximate solutions to this augmented objective, the inequalities~(\ref{eq:bound1-new}-\ref{eq:bound2-new}) will also remain true due to Propositions~\ref{prop:mi-bound} and~\ref{prop:kl-bound}.

@@

In the neural networks literature, the most commonly used method using variational bounds similar to Equation~(\ref{eq:bound}) is the Variational Autoencoder \citep{kingma-auto-encoding-2014, bowman-generating-2016}, which corresponds to the setting where $\lambda=1$ and the predicted output is equal to the observed input.
The $\beta$-VAE \citep{higgins2017beta}, a variant of the Variational Autoencoder, uses $\lambda > 1$ (whereas the Predictive Rate-Distortion objective~(\ref{eq:ib}) uses $\lambda \in (0,1)$), and has been linked to the Information Bottleneck by \cite{burgess2018understanding}.\footnote{As discussed in Section~\ref{sec:background}, our model becomes degenerate with $\lambda \ge 1$. Surprisingly, $\beta$-VAE operates in exactly this parameter regime, although }

@@

\section{Estimating Predictive Rate-Distortion   for Natural Language}\label{sec:language}
